A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-03-13 crystal-gel/nature/Supplementary.tex
Description of the iterative process

@@

This theory relies on an equation of state (EoS) for hard spheres to deduce phase boundaries of the colloids+polymers system. Canonically, the Carnahan-Starling (CS) EoS is used for the fluid and the equation of state by Hall for the crystal. CS is accurate for a wide range of volume fraction, but lacks a divergence below 1 although a divergence of the pressure is expected at the jamming point (random close packing $\phi_J\approx 0.64$). However CS is accurate up to 0.56. Therefore CS (coupled will Hall EoS) always describes accurately fluid-crystal coexistence since the fluid volume fraction is at most 0.495.

The gas spinodal line is given by a local maximum in free energy, a local property of the equation of state. Therefore, the locally correct CS EoS is able to describe accurately the gas side of the spinodal region up to 0.56. However if the critical point is situated at a larger volume fraction, CS fails to locate it properly. In the same way, the liquid side of the spinodal as described by CS can reach unphysical volume fractions (above close packing). Even more annoying, the binodal line, given by equating pressures and chemical potentials in the gas and the liquid, is not a local property. Therefore, the position of the gas side of the binodal depends on the behaviour of the equation of state on the liquid side. Failure of CS at high volume fraction implies inaccurate binodal lines both gas and liquid.

Here, our small polymer/colloid size ratio pushes the gas-liquid critical point to high volume fractions, exemplifying the failure of CS EoS. Adding a divergence at Jamming into CS equation of state is not trivial. Attempts by Lefebre~\cite{LeFevre1972} and Kolafa~\cite{Kolafa2004} match CS at low density, have the right divergence close to jamming but fare less than CS at intermediate volume fractions (0.5-0.56), the exact range where our critical point lies. Here we decided to use an equation by Liu~\cite{Liu2006a} that is completely equivalent to CS up to 0.56 and then show good agreement with simulation data near jamming.

>>>>

\subsection*{Adjusting theoretical and experimental phase diagrams}

The volume fraction of a colloidal suspension is notoriously difficult to estimate experimentally~\cite{Poon2012}. It depends on the cube of the colloid diameter and is thus extremely sensitive to any error in size determination. Therefore here we determine the volume fraction directly from the phase behaviour and deduce a precise measure of the (effective) colloid diameter from it.

Our two constraints is (i) to have the final local volume fraction of the 12-neighboured particles in the crystallising sample to be equal to the theoretical equilibrium crystal volume fraction (ii) to have the theoretical spinodal line to correspond to the gel region boundary. Starting from our initial estimates of $\sigma$ and $R$ we iteratively converge to $R=80$~nm, $\sigma = 2.25\,\mu$m and a crystal at $0.723$.

@@

The free volume theory relies on an equation of state (EoS) for hard spheres to deduce phase boundaries of the colloids+polymers system. Canonically, the Carnahan-Starling (CS) EoS is used for the fluid and the equation of state by Hall for the crystal. CS is accurate for a wide range of volume fraction, but lacks a divergence below 1 although a divergence of the pressure is expected at the jamming point (random close packing $\phi_J\approx 0.64$). However CS is accurate up to 0.56. Therefore CS (coupled will Hall EoS) always describes accurately fluid-crystal coexistence since the fluid volume fraction is at most 0.495.

The gas spinodal line is given by a local maximum in free energy, a local property of the equation of state. Therefore, the locally correct CS EoS is able to describe accurately the gas side of the spinodal region up to 0.56. The iterative process described above can be done only using CS EoS.

However, the liquid side of the binodal and spinodal lines obtained from CS EoS reach unphysically high volume fractions (above close packing), making comparisons with the measured liquid compositions impossible.

Indeed our small polymer/colloid size ratio pushes the gas-liquid critical point to high volume fractions, exemplifying the failure of CS EoS. Adding a divergence at Jamming into CS equation of state is not trivial. Attempts by Lefebre~\cite{LeFevre1972} and Kolafa~\cite{Kolafa2004} match CS at low density, have the right divergence close to jamming but fare less than CS at intermediate volume fractions (0.5-0.56), the exact range where our critical point lies. Here we decided to use an equation by Liu~\cite{Liu2006a} that is completely equivalent to CS up to 0.56 and then show good agreement with simulation data near jamming.

The Liu EoS allows us to draw a phase diagram that is in qualitative agreement with the measured liquid compositions.
